---
title: "ST_2_spatial"
author: "Sean Swift"
date: "January 30, 2020"
output:
  html_document: default
  pdf_document: default
---

# Spatial analyses

## Pulling out spatial data from rasters
We have GPS coordinates for each sample. There are freely available raster data sets for environmental variables (elevation, rainfall, temp).
Using the sample coordinates, pull out these environmental variables and join them to the sample metadata
```{r import}
library(raster)
library(ggplot2)
library(dplyr)
library(rgdal)
library(geosphere)
library(vegan)
#read in data

ST_16S <- readRDS("data/processed/cleaned/ST_16S.rds")


## select just samples that were collected in the field and have lat long (no controls)

field_meta <- ST_16S$metadata %>% filter(!is.na(lat) & !is.na(long))


```

### Rainfall

```{r rainfall}
# rasters are yearly averages, and can be
# found at rainfall atlas of hawaii and evapotranspiration hawaii
# http://rainfall.geography.hawaii.edu/rainfall.html
#Extract data from rasters using sampling coordinates

set_proj <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84"


#use annual rainfall txt file to make raster

oahu_rainfall <- raster("data/raw/spatial/rfgrid_mm_oahu_ann.txt", crs = set_proj)
oahu_rainfall

# extract values
field_meta$rainfall <-  raster::extract(x = oahu_rainfall,
                        y = field_meta[c("long","lat")],
                        df = TRUE)[[2]]



#makes raster map of oahu
plot(oahu_rainfall, main="Annual Rainfall Oahu")
points(x=field_meta$long, y=field_meta$lat, pch=20, col="black", cex=1)
points(x=field_meta$long, y=field_meta$lat, pch=20, col="white", cex=0.4)


```

### Elevation

```{r elevation}
# source: http://www.soest.hawaii.edu/coasts/data/hawaii/dem.html
#  Code to correct project for raw raster. Takes a lot of resources to run. 
 oahu_elevation <- raster("data/raw/spatial/oahu_dem/w001001.adf")
 oahu_elevation <- projectRaster(oahu_elevation, crs = set_proj)
# saveRDS(oahu_elevation,"data/interim/oahu_elev_raster.RDS")

# oahu_elevation <- readRDS("data/interim/oahu_elev_raster.RDS")
oahu_elevation

# extract values
field_meta$elevation <-  raster::extract(x = oahu_elevation,
                        y = field_meta[c("long","lat")],
                        df = TRUE)[[2]]


plot(oahu_elevation)
points(x=field_meta$long, y=field_meta$lat, pch=20, col="black", cex=1)
points(x=field_meta$long, y=field_meta$lat, pch=20, col="white", cex=0.4)

```


### C-CAP

```{r}

# ccap <- raster("data/raw/spatial/CCAP/hi_oahu_2011_ccap_hr_land_cover20140619.img")
# 
# # need to change projection, but process is taking a long time
# 
# #ccap <- projectRaster(ccap, crs = set_proj)
# 
# ccap
# 
# 
# plot(ccap)

```

### Distance from shore for each site
Anthony suggests organizing the gradient by distance from shore, which varies colinearly with precipitation, elevation, etc.

```{r distance shore}
# Choose arbitrary point on shore

shore <- c(-158.063539,21.640629)

# Calculate distance for each sample

field_meta$shore_dist <- geosphere::distGeo(shore, field_meta[c("long","lat")])

# Show colinearity of rainfall, elevation, distance from shore


plot(
  field_meta$site_order,
  field_meta$shore_dist,
  col = "red",
  pch = 20,
  xlab = "Sampling Site",
  ylab = "Value",
  main = "Variation in Distance, Rainfall, and Elevation"
  )
  
  points(field_meta$site_order,
  field_meta$rainfall,
  col = "blue",
  pch = 20)
  
  points(field_meta$site_order,
  field_meta$elevation,
  col = "green",
  pch = 20)
  
  legend(
  x = 0,
  y = 10000,
  c("Distance from shore", "Rainfall", "Elevation"),
  col = c("red", "blue", "green"),
  pch = 20
  )

  # write to an r file
  saveRDS(field_meta,"data/interim/field_meta.rds")
```





