---
title: "SuperTransect - Organizing Microbe Data For MMVEC"
output:
  html_document:
    df_print: paged
---
### Purpose:
We have microbial abundance (species x samples) and microbial taxonomy as flat .csv files.
Most R statistical methods want samples (observations) as rows and species as columns (variables).

In order to use mmvec, we need to transpose the abundance table so that rows are species and columns are samples.

# Read in the data

Let's read in the microbial abundance, taxonomy, and the MS metadata sheet.
We want microbial samples and MS samples to corellate 1:1.
There are extra microbial samples that don't correspond to MS samples and vice versa. 
Those samples need to be dropped

 *Microbial Abundance Data*
Columns:

- label   = clustering threshold. nNot necessary (just let's you know data was not clustered by % similarity).

- Group   = unique ID for the microbial PCR products. This lets us join abundance to the metadata table

- otu00003, otu00004, etc.  = DNA sequence counts for each 'species' of microbe. The column names match the 'OTU' column the taxonomy table. 

```{r}
library(data.table)

setwd("/home/sean/Documents/Bioinformatics/Projects/SuperTransect/Analysis/exploration/R_notebook")


abundance <- fread("data/processed/ST_16S_abundance.tsv", sep = "\t", header = T, key = "Group")
abundance[1:5,1:5]
```

*Microbial Taxonomy Data*

Currently, each taxanomic rank is separated into columns. 
For mmvec, we want all columns mashed together with ";" separating each rank.

Columns:
- OTU   = unique identifier for each microbial 'species'. Matches the column names of the abundance data
- Size  = total sequences in this OTU. Not really needed.

-kingdom, phylum, etc.  = clasification of the otu at different taxonomic ranks 


```{r}
taxonomy <- fread("data/processed/ST_16S_taxonomy.tsv", sep = "\t", header = T, key = "OTU")
taxonomy[1:5,1:5]

```

*Metadata*

This should be familiar! Slightly different formatting from the MS data, but mostly the same information.

Columns:

- id              = unique id for each microbial PCR product that was sequenced. 
                  Not exactly the same as 'sample_id' because some samples were sequenced multiple times.
                  This column matches the abundance table column 'Group'.
                  
                  
sample_barcode  = Unique id for each sample taken out of the field.
MS_barcode      = Unique id for corresponding MS sample taken out of the field.
                  This is the one we care about for mmvec!

And lots of other useful sample metadata columns that we won't use....

```{r}
metadata <- fread("data/processed/ST_16S_metadata.tsv", header = T, sep = "\t", key = "id")

# These are the columns we want
important_columns <- c("id",
                       "sample_barcode",
                       "MS_barcode",
                       "sample_type")

# Subset metadata to just the important columns
metadata[ , ..important_columns]

```

# Fix Potential Problems

Some microbial samples don't have a corresponding MS sample. Let's get rid of those samples
Are all the MS barcodes unique? They should be...

```{r}
message("Number of rows in metadata")
nrow(metadata)

message("Number of rows that have MS data")
metadata_ms <- na.omit(metadata, cols = "MS_barcode")
nrow(metadata_ms)

message("Number of duplicated MS barcodes")
duplicated_ms <- metadata_ms[duplicated(MS_barcode, by = MS_barcode),]
nrow(duplicated_ms)

duplicated_ms

```

Okay, so we sequenced the ridge water samples twice. Let's drop one set of duplicates. 
Note, we're also missing the marine water and sediment samples for now.

```{r}
# drop extra water samples
samples_to_drop <- metadata_ms[transect_name == "STSummit" & grepl("Water-DNA",collection_label), id]

metadata_ms <- metadata_ms[!id %in% samples_to_drop,]
nrow(metadata_ms)

message("any duplicated samples left?")
anyDuplicated(metadata_ms, by = "MS_barcode")

```

## Make All The Data Match Up
Now that we have chosen the right samples, let's match up the abundance and taxnomomy tables to the MS barcodes.

```{r}
# subset abundance
abundance_ms <- abundance[Group %in% metadata_ms$id]

# subset taxnomy (just in case some OTUs no longer exist)
taxonomy_ms <- taxonomy[OTU %in% colnames(abundance)[3:ncol(abundance_ms)]]



```


## Transform Data For mmvec
For mmvec, we need to change the file format of the data.

Microbial Abundance:  Transpose so 'species' are rows and samples are columns. Write out as tsv, then conver to biom.
Taxonomy:             Concatenate all the taxanomic ranks with ";"
Metadata:             Make MS_barcode the main identifier for all samples

*Abundance File*
Fix names, transpose, write out as biom.

```{r}
## Abundance file fromatting
library(biomformat)

# pull out ids
seq_ids  <- abundance_ms[, Group]
otus_ids <- colnames(abundance_ms)[3:ncol(abundance_ms)]

# drop samples missing from abundance and match the order of abundance rows
metadata_ms <- metadata_ms[id %in% seq_ids]
metadata_ms <- metadata_ms[order(match( id, seq_ids))]

# MS barcodes will be the new seq_ids
new_seq_ids <- metadata_ms$MS_barcode

# It's now safe drop the group column and label column
abundance_ms[, c("Group","label"):=NULL]

# transpose
t_abundance_ms <- transpose(abundance_ms)

# assign MS_barcods as column names (i.e. sample names)
colnames(t_abundance_ms) <- as.character(new_seq_ids)

# assign otu names as rownames
row.names(t_abundance_ms) <- otus_ids

# write out as biom file (v1)
# write_biom(t_abundance_ms, "data/processed/cleaned/mmvec/transect_microbes_mmvec.biom")

# write out as csv file
fwrite(t_abundance_ms, "data/processed/cleaned/mmvec_microbe_data/transect_microbes_mmvec.csv", row.names = T)


t_abundance_ms[1:5,1:5]
```


*Taxonomy File*
Concatenate names and write out as tsv file

```{r}
# columns to concatenate
columns_to_smoosh <- colnames(taxonomy)[-c(1:2)]
columns_to_smoosh

# prefixes to add to each column
prefixes <- c("k__","p__", "c__","o__","f__","g__")

# paste prefixes onto each column
for (i  in seq_along(prefixes)) {
  

taxonomy_ms[ , (columns_to_smoosh[i]) := lapply(.SD, function(x, pref) paste(pref, x, sep = ""), pref=prefixes[i]), .SDcols = columns_to_smoosh[i]]
}
# smoosh columns 
taxonomy_ms[, Taxon:= do.call(paste, c(.SD, sep = ";")), .SDcols = columns_to_smoosh]

# select OTU and new taxonomy column
taxonomy_clean <- taxonomy_ms[, .(OTU, Taxon)]
taxonomy_clean$Taxon <- paste(taxonomy_clean$Taxon, "s__",sep = ";")

# rename OTU to Feature ID
setnames(taxonomy_clean, "OTU", "Feature ID")

# add prefixes to each taxonomic rank

fwrite(taxonomy_clean, "data/processed/cleaned/mmvec_microbe_data/microbe_taxonomy_mmvec.tsv", sep = "\t", row.names = F)

taxonomy_clean[1:5]
```


# All Done !

Abundance file is transposed. The column names match MS sample barcodes. The data is written out as a biom file.

Taxonomy file is concatenated and written out as a tsv file. Following mmvec github, the taxonomy is stored in column "Taxon". 
The OTU ids are the first column, which is called "OTU". Hopefully that's okay. 




