---
output:
  html_document:
    df_print: paged
---

# Rayshader Attempt

Here we'll try to make 3D maps for fun data visualization!

To start, install the "leaflet" package.

To get rayshader on a mac, you will need to download and install XQuartz (xquartz.org).
Then, install the "rgl" package. Then carry out these lines:

install.packages("devtools")
devtools::install_github("tylermorganwall/rayshader")

### read in points
 
 
```{r}
sample_data <- readRDS("data/processed/cleaned/ST_16S_spatial.rds")

coords <- sample_data$metadata[c("long","lat")]

plot(coords)
```
 


### Step 1 - Bounding Box

Create a bounding box for the desired area (Waimea). The leaflet package will display it on a map

```{r}
library(rayshader)
library(leaflet)
library(dplyr)

# define bounding box with longitude/latitude coordinates
bbox <- list(
  p1 = list(long = -158.086249, lat = 21.588450),
  p2 = list(long = -157.949029, lat = 21.662884)
)

leaflet() %>%
  addTiles() %>% 
  addRectangles(
    lng1 = bbox$p1$long, lat1 = bbox$p1$lat,
    lng2 = bbox$p2$long, lat2 = bbox$p2$lat,
    fillColor = "transparent"
  ) %>%
  fitBounds(
    lng1 = bbox$p1$long, lat1 = bbox$p1$lat,
    lng2 = bbox$p2$long, lat2 = bbox$p2$lat
  )
```

### Step 2 - Define Image Size

There are 2 requirements:
1. The image dimensions need to match our bounding box coordinates (i.e have the same aspect ratio)
2. The image dimensions of the elevation data and the map overlay need to match exactly (i.e. produce the same size arrays)

We are using custom functions from Will Bishop (wcmbishop on Github).

```{r}
#the function
define_image_size <- function(bbox, major_dim = 400) {
  # calculate aspect ration (width/height) from lat/long bounding box
  aspect_ratio <- abs((bbox$p1$long - bbox$p2$long) / (bbox$p1$lat - bbox$p2$lat))
  # define dimensions
  img_width <- ifelse(aspect_ratio > 1, major_dim, major_dim*aspect_ratio) %>% round()
  img_height <- ifelse(aspect_ratio < 1, major_dim, major_dim/aspect_ratio) %>% round()
  size_str <- paste(img_width, img_height, sep = ",")
  list(height = img_height, width = img_width, size = size_str)
}

#our data
image_size <- define_image_size(bbox, major_dim = 600)

```

### Step 3 - Download Elevation Data

Using another custom function, pull elevation data from the USGS National Map tool API

```{r}
#the function
get_usgs_elevation_data <- function(bbox, size = "400,400", file = NULL, 
                                    sr_bbox = 4326, sr_image = 4326) {
  require(httr)
  
  # TODO - validate inputs
  
  url <- parse_url("https://elevation.nationalmap.gov/arcgis/rest/services/3DEPElevation/ImageServer/exportImage")
  res <- GET(
    url, 
    query = list(
      bbox = paste(bbox$p1$long, bbox$p1$lat, bbox$p2$long, bbox$p2$lat,
                   sep = ","),
      bboxSR = sr_bbox,
      imageSR = sr_image,
      size = size,
      format = "tiff",
      pixelType = "F32",
      noDataInterpretation = "esriNoDataMatchAny",
      interpolation = "+RSP_BilinearInterpolation",
      f = "json"
    )
  )
  
  if (status_code(res) == 200) {
    body <- content(res, type = "application/json")
    # TODO - check that bbox values are correct
    # message(jsonlite::toJSON(body, auto_unbox = TRUE, pretty = TRUE))
    
    img_res <- GET(body$href)
    img_bin <- content(img_res, "raw")
    if (is.null(file)) 
      file <- tempfile("elev_matrix", fileext = ".tif")
    writeBin(img_bin, file)
    message(paste("image saved to file:", file))
  } else {
    warning(res)
  }
  invisible(file)
}

# download elevation data
elev_file <- file.path("data/raw/", "sf-elevation.tif")

# uncomment to re-download the file
#get_usgs_elevation_data(bbox, size = image_size$size, file = elev_file,
#                        sr_bbox = 4326, sr_image = 4326)

```

### Step 4 - Plot the Map!

First, read the elevation data (.tif file) using another function.
Then, use rayshader code to calculate data layers and build a 2D plot.

```{r}
# load elevation data
elev_img <- raster::raster(elev_file)

elev_matrix <- matrix(
  raster::extract(elev_img, raster::extent(elev_img), buffer = 1000), 
  nrow = ncol(elev_img), ncol = nrow(elev_img)
)

# calculate rayshader layers
ambmat <- ambient_shade(elev_matrix, zscale = 30)
raymat <- ray_shade(elev_matrix, zscale = 30, lambert = TRUE)
watermap <- detect_water(elev_matrix)

# plot 2D
elev_matrix %>%
  sphere_shade(texture = "imhof4") %>%
  add_water(watermap, color = "imhof4") %>%
  add_shadow(raymat, max_darken = 0.5) %>%
  add_shadow(ambmat, max_darken = 0.5) %>%
  plot_map()


```



### Step 6b - 3D Ma

Increasing alphalayer in add_overlay makes overlay darker (less transparent)
In plot_3d:
theta refers to left/right rotation about z axis
phi refers to how steep a view you get
fov is the field of view (default is isometric)

```{r}
# coordinates function
find_image_coordinates <- function(long, lat, bbox, image_width, image_height) {
  x_img <- round(image_width * (long - min(bbox$p1$long, bbox$p2$long)) / abs(bbox$p1$long - bbox$p2$long))
  y_img <- round(image_height * (lat - min(bbox$p1$lat, bbox$p2$lat)) / abs(bbox$p1$lat - bbox$p2$lat))
  list(x = x_img, y = y_img)
}

# define label



coords_list <- apply(coords,1, FUN = function(x){ 
                    pos <- find_image_coordinates(
                    long = x[1], lat = x[2] , bbox = bbox,
                    image_width = image_size$width, image_height = image_size$height)
                    return(pos)
                  }

)

zscale <- 10
rgl::clear3d()
elev_matrix %>%
  sphere_shade(texture = "imhof1") %>%
  add_water(watermap, color = "imhof1") %>%
  add_shadow(raymat, max_darken = 0.5) %>%
  add_shadow(ambmat, 0) %>%
  plot_3d(elev_matrix, zscale = zscale, windowsize = c(1200, 1000),
          water = TRUE, wateralpha = 0.5,
          theta = -120, phi = 20, zoom = 0.65, fov = 0)


render_snapshot()
```

# Try to add some points

```{r}

library(geoviz)
library(viridisLite)
# fixed function by adding raise_agl to alt 

# pull out sample points
samps <- sample_data$metadata[!is.na(sample_data$metadata$long) & !is.na(sample_data$metadata$transect_name),]
samps <- sample_data$metadata %>% group_by(long, lat, shore_dist) %>% summarise %>% filter(!is.na(shore_dist))


map_viridis <- function(vec, num) {

  vector_expanded <-round(vec, 1) * 10 # expand to allow for decimal precision
  vector_exp_range <- max(vector_expanded) - min(vector_expanded)

  colour_vector <- viridis(vector_exp_range + 1) # get vector of colour values for all possible decimals between min and max value
  value_to_colour <- colour_vector[num * 10 - min(vector_expanded) + 1] # retrieve colour value for number

  return(value_to_colour)

}


samps$colors <- sapply(samps$shore_dist, function(x) map_viridis(samps$shore_dist, x))

# this came from library geoviz
# "fixed" add_gps by cutting out all the line related items, and having it plot spheres instead of points
# also just made it so params get written out and I can rgl:: spheres3d separately
source("src/fixed_add_gps.R")

sites_spheres <- custom_add_spheres(  elev_img,
                                        samps$lat,
                                        samps$long,
                                        raise_agl = 80,
                                        zscale = zscale,
                                        colour = samps$colors,
                                        point_size = 5,
                                        rad = 2)

```
## check the image
```{r}
## still version for checking image quality




rgl::clear3d()

rgl::points3d(sites_spheres$x, sites_spheres$y, sites_spheres$z, 
                   color = sites_spheres$col, alpha = sites_spheres$alpha, size = sites_spheres$size)

#rgl::spheres3d(sites_spheres$x, sites_spheres$y, sites_spheres$z, 
 #                  color = sites_spheres$col, alpha = sites_spheres$alpha, size = sites_spheres$size, radius = sites_spheres$radius)
elev_matrix %>%
  sphere_shade(texture = "imhof1") %>%
  add_water(watermap, color = "desert") %>%
  add_shadow(raymat, max_darken = 0.5) %>%
  add_shadow(ambmat, 0) %>%
  rayshader::plot_3d(
    elev_matrix,
    zscale = zscale,
    windowsize = c(1200, 1000),
    water = F,
    wateralpha = 0.2,
    theta = 237,
    phi = 25,
    zoom = 0.55,
    fov = 0,
    shadow = T
    )
rgl::points3d(sites_spheres$x, sites_spheres$y, sites_spheres$z, 
                   color = sites_spheres$col, alpha = sites_spheres$alpha, size = sites_spheres$size)
render_snapshot()
#rgl.postscript("tmp.txt", fmt = "svg")
# snapshot3d("tmp.png")
 
# 
#  add_gps_to_rayshader(
  # elev_img,
  # samps$lat,
  # samps$long,
  # raise_agl = 50,
  # line_width = 1,
  # lightsaber = F,
  # zscale = zscale,
  # ground_shadow = F,
  # colour = samps$colors,
  # clamp_to_ground = T,
  # as_line = F,
  # point_size = 10,
  # rad = 2
# )
# 
# render_snapshot()
```
## MAKE A GIF OMG


```{r}
# from https://wcmbishop.github.io/rayshader-demo/#one-can-simply-gif
# which is sooo goood, I pulled the function save_3d_gif
# "fixed" by adding lines 77:78 to allow adding spheres to the plot for points
source("src/rayshader_demo/rayshader-gif.R")



sites_spheres <- custom_add_spheres(  elev_img,
                                        samps$lat,
                                        samps$long,
                                        raise_agl = 80,
                                        zscale = zscale,
                                        colour = samps$colors,
                                         point_size = 3,
                                        rad = 2)

n_frames <- 180
thetas <- transition_values(from = 230, to = 280, steps = n_frames)
# generate gif
zscale <- 10
base_shade <- elev_matrix %>% 
                sphere_shade(texture = "imhof1") %>% 
                add_water(watermap, color = "imhof1") %>%
                add_shadow(raymat, max_darken = 0.5) %>%
                add_shadow(ambmat, 0)
  
  




  save_3d_gif(hillshade = base_shade, elev_matrix, file = "outputs/test4.gif", duration = 9,
              solid = TRUE, shadow = TRUE, water = TRUE, zscale = zscale,
              watercolor = "imhof1", wateralpha = 0.2,
              waterlinecolor = "#ffffff", waterlinealpha = 0.2,
              theta = thetas, phi = 25, sites_spheres = sites_spheres)
```

